\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-17 submission/manuscript.Rmd
update bibs

Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Most notably, there are misleading practices of using "leaky" ML pipelines where models are tested on training data or reporting the best outcome of different iterations of cross-validation. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

>>>>

Begüm D. Topçuoğlu${^1}$, Nick Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex and uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016;  . More notably, there are flawed ML practices like using ML pipelines where there is no seperate held-out test dataset to evaulate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

@@

#### Interpretation of results
 Our results show that 
 Suprisingly, l2 logistic..
 computational efficiency and interp of logistic vs random forest
 decision tree and l1 why so low?

#### Consideration of possible weaknesses
 What happens with imbalanced data
 What happens with smaller datasets
 Does the code work on any microbiome data?
  Why did we use only one dataset? Becuase we are comparing several methods on 1 dataset.
  What will the future data look like?

#### Relationship of results to previous literature and broader implications of having answered research question

#### Prospects for future progress
  Subsample dataset and refit see how method behaves for 7 methods. (500->250->50)

### 2019-05-17 submission/manuscript.tex
update bibs

Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Mack Ruffin\({^3}\),
Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
However, currently the field's use of ML lacks clarity and consistency
on which methods are used and how these methods are implemented. Most
notably, there are misleading practices of using ``leaky'' ML pipelines
where models are tested on training data or reporting the best outcome
of different iterations of cross-validation. Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11--14).
The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and reliable machine learning pipelines; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase reliable ML pipeline and to shed light on how much ML model
selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite the simplicity, the
L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied only 0.021, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

@@

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (15). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic health was labeled by colonoscopy with adequate preparation and
tissue histopathology of all resected lesions. Patients with an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology were classified as advanced adenoma. Study had
172 patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(16--18).

@@

\hypertarget{ref-galkin_human_2018}{}
12. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
13. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
14. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.

\hypertarget{ref-sze_leveraging_2018}{}
15. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
16. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-westcott_opticlust_2017}{}
17. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-rognes_vsearch_2016}{}
18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

>>>>

Begüm D. Topçuoğlu\({^1}\), Nick Lesniak\({^1}\), Jenna Wiens\({^2}\),
Mack Ruffin\({^3}\), Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented {[}(12); . Most notably, there are misleading practices of
using ``leaky'' ML pipelines where models are tested on training data,
reporting few or only the best outcomes of different randomizations of
cross-validation and showing a disregard for large differences between
cross-validation and testing performances as well as large confidence
intervals of testing performances (4, 13--18). Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11,
19--21). The lack of transparency on modeling methodology and model
selection negatively impact model reproducibility and reliability. We
need to strive toward better machine learning practices by (1)
implementing consistent and reliable machine learning pipelines; (2)
selecting ML models that reflect the goal of the study as it will inform
our expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization and prediction performance of each ML model.
The median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite the simplicity, the
L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied only 0.021, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

\paragraph{Interpretation of results}\label{interpretation-of-results}

Our results show that Suprisingly, l2 logistic.. computational
efficiency and interp of logistic vs random forest decision tree and l1
why so low?

\paragraph{Consideration of possible
weaknesses}\label{consideration-of-possible-weaknesses}

What happens with imbalanced data What happens with smaller datasets
Does the code work on any microbiome data? Why did we use only one
dataset? Becuase we are comparing several methods on 1 dataset. What
will the future data look like?

\paragraph{Relationship of results to previous literature and broader
implications of having answered research
question}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-having-answered-research-question}

\paragraph{Prospects for future
progress}\label{prospects-for-future-progress}

Subsample dataset and refit see how method behaves for 7 methods.
(500-\textgreater{}250-\textgreater{}50)

@@

The data used for this analysis are stool bacterial abundances and
clinical information of the patients recruited by Great Lakes-New
England Early Detection Research Network study. These data were obtained
from Sze et al (22). The stool samples were provided by recruited adult
participants who were undergoing scheduled screening or surveillance
colonoscopy. Colonoscopies were performed and fecal samples were
collected from participants in four locations: Toronto (ON, Canada),
Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
colonic health was labeled by colonoscopy with adequate preparation and
tissue histopathology of all resected lesions. Patients with an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology were classified as advanced adenoma. Study had
172 patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(23--25).

@@

\hypertarget{ref-thaiss_persistent_2016}{}
12. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
alterations modulate the rate of post-dieting weight regain. Nature
\textbf{540}:544--551.
doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.

\hypertarget{ref-flemer_oral_2018}{}
13. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
2018. The oral microbiota in colorectal cancer is distinctive and
predictive. Gut \textbf{67}:1454--1463.
doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.

\hypertarget{ref-dai_multi-cohort_2018}{}
14. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
identified altered bacteria across populations and universal bacterial
markers. Microbiome \textbf{6}:70.
doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.

\hypertarget{ref-montassier_pretreatment_2016}{}
15. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
2016. Pretreatment gut microbiome predicts chemotherapy-related
bloodstream infection. Genome Medicine \textbf{8}:49.
doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.

\hypertarget{ref-papa_non-invasive_2012}{}
16. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
\textbf{Weber S}, \textbf{Preheim SP}, \textbf{Gevers D},
\textbf{Giannoukos G}, \textbf{Ciulla D}, \textbf{Tabbaa D},
\textbf{Ingram J}, \textbf{Schauer DB}, \textbf{Ward DV},
\textbf{Korzenik JR}, \textbf{Xavier RJ}, \textbf{Bousvaros A},
\textbf{Alm EJ}. 2012. Non-invasive mapping of the gastrointestinal
microbiota identifies children with inflammatory bowel disease. PLOS ONE
\textbf{7}:e39242.
doi:\href{https://doi.org/10.1371/journal.pone.0039242}{10.1371/journal.pone.0039242}.

\hypertarget{ref-mossotto_classification_2017}{}
17. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
Classification of paediatric inflammatory bowel disease using machine
learning. Scientific Reports \textbf{7}.
doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.

\hypertarget{ref-ai_systematic_2017}{}
18. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
supervised classifiers for fecal microbiota-based prediction of
colorectal cancer. Oncotarget \textbf{8}:9546--9556.
doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.

\hypertarget{ref-galkin_human_2018}{}
19. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
2018. Human microbiome aging clocks based on deep learning and tandem of
permutation feature importance and accumulated local effects. bioRxiv.
doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.

\hypertarget{ref-reiman_using_2017}{}
20. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
convolutional neural networks to explore the microbiome, pp. 4269--4272.
\emph{In} 2017 39th annual international conference of the IEEE
engineering in medicine and biology society (EMBC).

\hypertarget{ref-fioravanti_phylogenetic_2017}{}
21. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.

\hypertarget{ref-sze_leveraging_2018}{}
22. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-westcott_opticlust_2017}{}
24. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-rognes_vsearch_2016}{}
25. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
